---
title: "Using R for regional optimizaiton of runout with the GPP tool in SAGA-GIS"
author: "Jason Goetz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Regional runout grid search optimization}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
comment = "#>")
knitr::opts_knit$set(root.dir = "/home/jason/Data/Chile/")
```

The runout.opt package was developed for automatic parameter selection for
single event and regional runout modeling using the random walk and PCM components of the [Gravitational 
Process Path (GPP) Model](https://gmd.copernicus.org/articles/10/3309/2017/) tool in [SAGA-GIS](http://www.saga-gis.org). The optimization procedure uses a two-stage approach,
where we first optimize the random walk model to find the 'best' simulation of runout path,
and then plug-in these values to the PCM model to optimize for runout distance. The performance
of the runout path is based on the area under the receiver operating characteristic curve (AUROC),
and runout distance is based on a measure of relative error.

This vignette

* Provides an example of regionally mapping debris flow runout 
* Shows how to setup and run grid search for regional optimization
* Demonstrates the use of parallel computing for faster results
* Illustrates the use of spatial cross-validation to validate model performance


### Load packages and data

To start, we will load all the necessary packages and data. The `runout.opt` functions
are made to work with `RasterLayer`, `SpatialPointsDataFrame` and `SpatialPolygonsDataFrame`
objects from the `raster` and `sp` packages. 

We are using an 12.5 m spatial resolution DEM. The debris-flow
runout track and source point that we will use for runout model optimization are one 
of many stored in shapefile. So, we select a single runout track, and the corresponding 
source point for this example.

```{r, fig.height=6, fig.width = 6}
library(runout.opt)
library(raster)
library(rgdal)
library(sp)
library(Rsagacmd)

# Load digital elevation model (DEM)
dem <- raster("elev_alos_12_5m.tif")

# Load runout source points
source_points <- readOGR(".", "debris_flow_source_points")

# Load runout track polygons and assign object ID based on row number
runout_polygons <- readOGR(".", "debris_flow_polys_sample")
runout_polygons$objectid <- 1:length(runout_polygons)

plot(dem)
plot(runout_polygons, add=TRUE)
```

### Initiate a SAGA-GIS geoprocessor object

The runout simulation is computed in SAGA-GIS. To do this, we are coupling
R with SAGA-GIS using the [`Rsagacmd`](https://github.com/stevenpawley/Rsagacmd) package.
Each time we load this package we need to initiate a SAGA-GIS geoprocessor object, 
which generates functions in R to SAGA-GIS libraries and tools. We only need to load the
Geomorphology library (i.e., sim_geomorphology) to access the GPP tool. This is also
faster than loading all of the SAGA-GIS libraries (e.g., `saga_gis()`)

```{r}
saga <- saga_gis(opt_lib = "sim_geomorphology")
```

### Regional GPP random walk grid search optimziation (runout path)

In our first stage of optimization, we will set up vectors to define our grid 
search space for the random walk runout path model component. This is an exhaustive
list of the parameters that will be tested.

```{r}
steps <- 11
rwexp_vec <- seq(1.3, 3, len=steps)
rwper_vec <- seq(1.5, 2, len=steps)
rwslp_vec <- seq(20, 40, len=steps)

rwexp_vec
rwper_vec
rwslp_vec

```

We will use parallelization to speed up our computations using the `foreach` package. This loop
will compute performances for all possible parameter combinations for each runout polygon
and store them as a list. In our case, this is our `rw_grid_search_multi` object. Depending on
the number of runout events and the grid search space size, this computation can take some time.
The `rwGridsearch` function has an option `save_res = TRUE` that allows us to save the grid search
results for each runout individually. This is useful in the case where processing fails since it avoids
the need to re-run the grid search for all runout events.

```{r, eval=FALSE}
library(foreach)

# Define which runout polygons are used for optimization
polyid_vec <- 1:100

# Set up cluster
cl <- parallel::makeCluster(7)
doParallel::registerDoParallel(cl)

# Run grid search loop
rw_grisearch_multi <-
  foreach(poly_id=polyid_vec, .packages=c('rgdal','raster', 'rgeos', 'ROCR', 'Rsagacmd', 'sf', 'runout.opt')) %dopar% {

    .GlobalEnv$saga <- saga

    rwGridsearch(dem, slide_plys = runout_polygons, slide_src = source_points,
                   slide_id = poly_id, slp_v = rwslp_vec, ex_v = rwexp_vec, per_v = rwper_vec,
                   gpp_iter = 1000, buffer_ext = 500, buffer_source = 50, save_res = FALSE,
                   plot_eval = FALSE)

  }

parallel::stopCluster(cl)
```

```{r, echo=FALSE}
setwd("/home/jason/Scratch/GPP_RW_Paper")
(load("rw_gridsearch_multi.Rd"))
```

To find the optimal parameter set, we will aggregate the performance values
of each runout event across grid search space using the median value.

```{r}
rw_opt <- rwGetOpt(rw_gridsearch_multi, measure = median)
rw_opt
```

### Validate random walk parameters' transferability using spatial cross validation 

We can perform spatial cross-validation to test the transferability of optimal parameter sets.
This is done using the k-means partitioning approach from the [`sperrorest`](https://github.com/cran/sperrorest)
package. In our example, we will perform 5-fold spatial cross-validation with 10 repetitions.

Additionally, we can visualize (`plot_freq = TRUE`) the relative frequency of optimal parameter sets from
all spatial cross-validation iterations.

```{r, fig.height=4, fig.width = 6}
rw_spcv <- rwSPCV(x = rw_gridsearch_multi, slide_plys = runout_polygons,
                  n_folds = 5, repetitions = 10)

freq_rw <- rwPoolSPCV(rw_spcv, plot_freq = TRUE)
freq_rw
```

### Regional GPP PCM model grid search optimzation (runout distance)

We can now plug-in the random walk optimal parameter set, and optimize for
runout distance using the PCM model component.

```{r}
# Define PCM model grid seach space
pcmmd_vec <- seq(20, 150, by=5) # mass-to-drag ratio (m)
pcmmu_vec <- seq(0.04, 0.6, by=0.01) # sliding friction coefficient 

pcmmd_vec
pcmmu_vec
```

```{r, eval=FALSE}
# Run using parallelization
cl <- parallel::makeCluster(4)
doParallel::registerDoParallel(cl)

pcm_gridsearch_multi <-
  foreach(poly_id=polyid_vec, .packages=c('rgdal','raster', 'rgeos', 'ROCR', 'Rsagacmd', 'sf', 'runout.opt')) %dopar% {

    .GlobalEnv$saga <- saga

    pcmGridsearch(dem,
                  slide_plys = runout_polygons, slide_src = source_points, slide_id = poly_id,
                  rw_slp = rw_opt$rw_slp_opt, rw_ex = rw_opt$rw_exp_opt, rw_per = rw_opt$rw_per_opt,
                  pcm_mu_v = pcmmu_vec, pcm_md_v = pcmmd_vec,
                  gpp_iter = 1000,
                  buffer_ext = 500, buffer_source = NULL,
                  predict_threshold = 0.5,
                  plot_eval = FALSE)

  }

parallel::stopCluster(cl)
```

```{r, echo=FALSE}
setwd("/home/jason/Scratch/GPP_PCM_Paper")
(load("pcm_gridsearch_multi.Rd"))
```

Next, we apply the `pcmGetOpt` function to find the regionally optimal PCM model parameters.
In our example, we are looking for the parameter set that results in the lowest median 
relative error across all runout events. The median model performances across grid search
space can be visualized using `plot_opt=TRUE`.

```{r, fig.height=4, fig.width = 6}
pcmGetOpt(pcm_gridsearch_multi, performance = "relerr", measure = "median", plot_opt = TRUE)
```

### Validate PCM model parameters' transferability using spatial cross validation 

Similar to the random walk model, we can explore the spatial transferability
of optimal parameters for the PCM model, and visualize the results.

```{r, fig.height=4, fig.width = 6}
pcm_spcv <- pcmSPCV(pcm_gridsearch_multi, slide_plys = runout_polygons,
                    n_folds = 5, repetitions = 10, from_save = FALSE)

freq_pcm <- pcmPoolSPCV(pcm_spcv, plot_freq = TRUE)
freq_pcm
```
