---
title: "Optimizing for a an individual runout event"
author: "Jason Goetz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Optimizing for a an individual runout event}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
comment = "#>")
knitr::opts_knit$set(root.dir = "/home/jason/Data/Chile/")
```

The runout.opt package was developed for automatic parameter selection for
single event and regional runout modeling using the random walk and PCM components of the [Gravitational 
Process Path (GPP) Model](https://gmd.copernicus.org/articles/10/3309/2017/) tool in [SAGA-GIS](http://www.saga-gis.org). The optimization procedure uses a two-stage approach,
where we first optimize the random walk model to find the 'best' simulation of runout path,
and then plug-in these values to the PCM model to optimize for runout distance. The performance
of the runout path is based on the area under the receiver operating characteristic curve (AUROC),
and runout distance is based on a measure of relative error.

This vignette provides an example of applying the runout.opt package for
selecting optimal parameter sets for a single debris flow event. It also,

* Shows how to measure and visualize GPP runout model performance in R
* Shows how to setup and run grid search optimization

### Load packages and data

To start, we will load all the necessary packages and data. The `runout.opt` functions
are made to work with `RasterLayer`, `SpatialPointsDataFrame` and `SpatialPolygonsDataFrame`
objects from the `raster` and `sp` packages. 

We are using an 12.5 m spatial resolution DEM. The debris-flow
runout track and source point that we will use for runout model optimization are one 
of many stored in shapefile. So, we select a single runout track, and the corresponding 
source point for this example.

```{r}
library(runout.opt)
library(raster)
library(rgdal)
library(sp)
library(Rsagacmd)

# Load digital elevation model (DEM)
dem <- raster("elev_alos_12_5m.tif")

# Load runout source points
source_points <- readOGR(".", "debris_flow_source_points")

# Load runout track polygons and assign object ID based on row number
runout_polygons <- readOGR(".", "debris_flow_polys_sample")
runout_polygons$objectid <- 1:length(runout_polygons)

# Select a debris flow and source point for this example
runout_polygon <- runout_polygons[77,]

sel_source_point  <- over(source_points, runout_polygon)
source_point <- source_points[!is.na(sel_source_point$objectid),]

plot(runout_polygon)
plot(source_point, add = TRUE)

```

### Initiate a SAGA-GIS geoprocessor object

The runout simulation is computed in SAGA-GIS. To do this, we are coupling
R with SAGA-GIS using the [`Rsagacmd`](https://github.com/stevenpawley/Rsagacmd) package.
Each time we load this package we need to initiate a SAGA-GIS geoprocessor object, 
which generates functions in R to SAGA-GIS libraries and tools. We only need to load the
Geomorphology library (i.e., sim_geomorphology) to access the GPP tool. This is also
faster than loading all of the SAGA-GIS libraries (e.g., `saga_gis()`)

```{r}
saga <- saga_gis(opt_lib = "sim_geomorphology")
```

### GPP random walk simulation (runout path)

We simulate the runout path using a random walk. The `rwPerformance` function does
this for us, as well as computing its performance by computing the AUROC based on
a comparison of the simulated and observed (mapped) runout paths. 

In our example, only a single point was mapped representing the source point.
To obtain a source area, we simply apply a buffer around the source point that
is clipped at the runout polygon boundaries. 

Note: The source of the runout ('slide_src') can be a single point, a buffered point
or a polygon representing a mapped source area.

```{r, fig.height=6, fig.width = 6}
rwPerformance(dem, slide_plys = runout_polygon, slide_src = source_point,
                   slp = 30, ex = 3, per = 2,
                   gpp_iter = 1000, buffer_ext = 500, buffer_source = 50,
                   plot_eval = TRUE, saga_lib = saga)
```

### GPP random walk grid search optimization

To test the random walk model under a variety of parameter sets, we need
to first define our grid search space. To do this we create vectors for each of
the random walk model parameters that include the parameter values we want test.

```{r}
steps <- 3
# Exponent controlling lateral spread
rwexp_vec <- seq(1.3, 3, len=steps)
# Persistence factor to weight flow direction consistency
rwper_vec <- seq(1.5, 2, len=steps)
# Slope threshold - below lateral spreading is modeled.
rwslp_vec <- seq(20, 40, len=steps)
```

We can compute performances values (i.e. AUROC) across our grid search
space.

```{r}
rw_gridsearch <- rwGridsearch(dem, slide_plys = runout_polygon, slide_src = source_point,
               #Input random walk grid search space
               slp_v = rwslp_vec, ex_v = rwexp_vec, per_v = rwper_vec,
               #Set number of simulation iterations
               gpp_iter = 1000,
               #Define processing extent size (m)
               buffer_ext = 500,
               #(Optional) Define size of buffer to make source area from point
               buffer_source = 50,
               saga_lib = saga)

rw_gridsearch
```

And find the parameter set that resulted in the highest AUROC score.

```{r}
rw_opt_single <- rwGetOpt_single(rw_gridsearch)
rw_opt_single
```

### GPP PCM model simulation (runout distance)

Now that we have obtained an optimal parameter set for the runout walk model,
we can plug-in these values to the PCM model that controls runout distance.

```{r, fig.height=6, fig.width = 6}
pcm <- pcmPerformance(dem, slide_plys = runout_polygon, slide_src = source_point,
               rw_slp = 40, rw_ex = 3, rw_per = 1.5,
               pcm_mu = 0.15, pcm_md = 120,
               gpp_iter = 1000, buffer_ext = 500, buffer_source = 50,
               plot_eval = TRUE, return_features = TRUE, saga_lib = saga)

# Runout distance relative error
pcm$length.relerr

# Plot GPP PCM runout modelling ouputs
gpp_output <- stack(pcm$gpp.parea, pcm$gpp.stop, pcm$gpp.maxvel)
names(gpp_output) <- c("Process_area", "Stop_positions", "Max_velocity")
plot(gpp_output)
```


### GPP PCM grid search optimization

Similar to the random walk optimization, we set up vectors to create our
grid search space, measure the runout distance model performance (e.g. using
relative error), and then find the parameter set with the best performance.

```{r}
# The mass-to-drag ratio (m)
pcmmd_vec <- seq(20, 120, by=20)
# The sliding friction coefficient
pcmmu_vec <- seq(0.05, 0.3, by=0.1)

pcm_gridsearch <- pcmGridsearch(dem,
                       slide_plys = runout_polygon, slide_src = source_point,
                       #Plug-in random walk optimal parameters
                       rw_slp = rw_opt_single$rw_slp_opt,
                       rw_ex = rw_opt_single$rw_exp_opt,
                       rw_per = rw_opt_single$rw_per_opt,
                       #Input PCM grid search space
                       pcm_mu_v = pcmmu_vec,
                       pcm_md_v = pcmmd_vec,
                       #Set number of simulation iterations
                       gpp_iter = 1000,
                       #Define processing extent size (m)
                       buffer_ext = 500,
                       #(Optional) Define size of buffer to make source area from point
                       buffer_source = 50, 
                       saga_lib = saga)

# Get optimal parameters
pcmGetOpt_single(pcm_gridsearch)

```
